The Calculus of Linear Constructions

### 2021-11-27 writing/draft4/lics22.tex
draft4

% packages
\usepackage{graphicx}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{mdwtab}
\usepackage{syntax}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{mathpartir}
\usepackage{listings}
\usepackage{lstlangcoq}
\usepackage{float}
\usepackage{tikz-cd}

\lstset{basicstyle=\small,language=Coq,showstringspaces=false}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

\renewcommand{\syntleft}{}
\renewcommand{\syntright}{}
\setlength{\grammarparsep}{7pt}
\setlength{\grammarindent}{7em}
\newcommand{\indalt}[1][2]{\\\hspace*{-1.2em}\textbar\quad}
\newcommand{\rname}[1]{\textsc{\footnotesize #1}}
\newcommand{\pure}[1]{|#1|}
\newcommand{\sort}{\textbf{sort}}
\newcommand{\type}{\textbf{type}}
\newcommand{\refl}{\text{refl}}
\newcommand{\letin}[3]{\ $\text{let }#1\text{ := }#2\text{ in }#3$\ }
\newcommand{\ind}{\ $\text{ind}$\ }
\newcommand{\iter}{\ $\text{iter}$\ }
\newcommand{\unit}{\text{unit}}
\newcommand{\new}{\textbf{new}}
\newcommand{\free}{\textbf{free}}
\newcommand{\get}{\textbf{get}}
\newcommand{\set}{\textbf{set}}
\newcommand{\utype}{:_{\scriptscriptstyle U}}
\newcommand{\ltype}{:_{\scriptscriptstyle L}}
\newcommand{\stype}[1]{:_#1}
\newcommand{\step}{\leadsto}
\newcommand{\pstep}{\leadsto}
\newcommand{\mrg}[3]{#1\ddagger#2\ddagger#3}
\newcommand{\erase}[1]{\llbracket #1 \rrbracket}
\newcommand{\lift}[1]{\llparenthesis #1 \rrparenthesis}

@@

  \section{Applications}
  In this section, we demonstrate some of the possible applications of CLC through examples. We collapse the sort annotations on arrows $(x \utype A) \rightarrow B$ and $(x \ltype A) \rightarrow B$ into $(x : A) \rightarrow B$ whenever $A$'s sort is inferable from context.

  \subsection{Stateful Reasoning}
  We first postulate five state axioms. These axioms could be viewed as an interface to a trusted memory allocator.
  \begin{definition}
    State axioms.
    \begin{itemize}
      \item $l \mapsto_A m : \mathbb{N} \rightarrow (A : U_k) \rightarrow A \rightarrow L_k$ \\
      The $\mapsto$ is a linear type constructor. It fulfills a role similar to L3's \cite{l3} memory access capability, or Separation Logic's \cite{reynolds02} points-to assertion. Intuitively, it is a proof that a term $m$ of non-linear type $A$ is currently stored at address $l$.
      \item $\new : (A : U_k) \rightarrow (m : A) \rightarrow F l:\mathbb{N}.(l \mapsto_A m)$ \\
      $\new$ is a function that allocates a new memory cell for term $m$ of non-linear type $A$. It results in an address $l$ paired with an assertion $(l \mapsto_A m)$.
      \item $\free : (A : U_k) \rightarrow F l:\mathbb{N}.F m:A.(l \mapsto_A m) \rightarrow \mathbf{1}$ \\
      $\free$ is a function that de-allocates the memory cell at location $l$ which is currently in use. This can be seen in its input $F l:\mathbb{N}.F m:A.(l \mapsto_A m)$, where the assertion component $(l \mapsto_A m)$ is a proof that location $l$ is currently allocated.
      \item $\get : (A : U_k) \rightarrow (l : \mathbb{N}) \rightarrow (m : A) \rightarrow (l \mapsto_A m) \rightarrow F \_: (\Sigma x. x =_A m). (l \mapsto_A m)$ \\
      $\get$ is a function that retrieves the term stored at address $l$ when given a proof $(l \mapsto_A m)$ that there is indeed a term $m$ currently stored there. The type of $\get$'s output is the rather complicated $F \_: (\Sigma x. x =_A m). (l \mapsto_A m)$. This not only returns the term stored at $l$, but also a proof that the returned term is exactly the same as what $(l \mapsto_A m)$ claimed to have stored. The proof $(l \mapsto_A m)$ is returned unchanged as well.
      \item $\set : (A\ B : U_k) \rightarrow (l : \mathbb{N}) \rightarrow (m : A) \rightarrow (l \mapsto_A m) \multimap (n : B) \multimap (l \mapsto_B n)$ \\
      $\set$ has the most complicated type of all the axioms presented here. Intuitively, if address $l$ is allocated and storing some term $m$ of type $A$, $\set$ may put term $n$ of type $B$ into the the store at address $l$, overwriting the original term. This is the so called "strong update" operation. The consumption of $(l \mapsto_A m)$ and return of $(l \mapsto_B n)$ accurately characterize this update process.
    \end{itemize}
  \end{definition}
  Due to the fact that $(l \mapsto_A m)$ is a linear type, proofs of this type can not be duplicated and may only be consumed by other state axioms. This is the same non-sharing principle that empowers Separation Logic. Unlike Separation Logic, $(l \mapsto_A m)$ itself is a first-class object in CLC, meaning that it can be the result of computation (large eliminations), input to computations, stored in data structures, etc.

@@

>>>>

% packages
\usepackage{graphicx}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{mdwtab}
\usepackage{syntax}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{mathpartir}
\usepackage{listings}
\usepackage{lstlangcoq}
\usepackage{float}
\usepackage{tikz-cd}
\usepackage{enumitem}

\lstset{basicstyle=\small,language=Coq,showstringspaces=false}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\setlist[itemize]{leftmargin=1.6em}
\renewcommand{\syntleft}{}
\renewcommand{\syntright}{}
\setlength{\grammarparsep}{7pt}
\setlength{\grammarindent}{7em}
\newcommand{\indalt}[1][2]{\\\hspace*{-1.2em}\textbar\quad}
\newcommand{\rname}[1]{\textsc{\footnotesize #1}}
\newcommand{\pure}[1]{|#1|}
\newcommand{\refl}{\text{refl}}
\newcommand{\letin}[3]{\ $\text{let }#1\text{ := }#2\text{ in }#3$\ }
\newcommand{\ind}{\ $\text{ind}$\ }
\newcommand{\iter}{\ $\text{iter}$\ }
\newcommand{\unit}{\text{unit}}
\newcommand{\new}{\text{new}}
\newcommand{\free}{\text{free}}
\newcommand{\get}{\text{get}}
\newcommand{\set}{\text{set}}
\newcommand{\session}{\text{session}}
\newcommand{\channel}{\text{channel}}
\newcommand{\open}{\text{open}}
\newcommand{\close}{\text{close}}
\newcommand{\send}{\text{send}}
\newcommand{\recv}{\text{recv}}
\newcommand{\SEND}{\texttt{SEND}}
\newcommand{\RECV}{\texttt{RECV}}
\newcommand{\END}{\texttt{END}}
\newcommand{\utype}{:_{\scriptscriptstyle U}}
\newcommand{\ltype}{:_{\scriptscriptstyle L}}
\newcommand{\stype}[1]{:_#1}
\newcommand{\step}{\leadsto}
\newcommand{\pstep}{\leadsto}
\newcommand{\mrg}[3]{#1\ddagger#2\ddagger#3}
\newcommand{\erase}[1]{\llbracket #1 \rrbracket}
\newcommand{\lift}[1]{\llparenthesis #1 \rrparenthesis}

@@

  \section{Applications}
  In this section, we demonstrate some of the possible applications of CLC through examples. We collapse the sort annotations on arrows $(x \utype A) \rightarrow B$ and $(x \ltype A) \rightarrow B$ into $(x : A) \rightarrow B$ whenever $A$'s sort is inferable from context. We also omit writing universe levels for the sake of clarity.

  \subsection{Stateful Reasoning}
  We first postulate five state axioms. These axioms could be viewed as an interface to a trusted memory allocator.
  \begin{definition}
    State axioms.
    \begin{itemize}
      \item $l \mapsto_A m : \mathbb{N} \rightarrow (A : U) \rightarrow A \rightarrow L$ \\
      The $\mapsto$ is a linear type constructor. It fulfills a role similar to L3's \cite{l3} memory access capability, or Separation Logic's \cite{reynolds02} points-to assertion. Intuitively, it is a proof that a term $m$ of non-linear type $A$ is currently stored at address $l$.
      \item $\textbf{new} : (A : U) \rightarrow (m : A) \rightarrow F l:\mathbb{N}.(l \mapsto_A m)$ \\
      $\new$ is a function that allocates a new memory cell for term $m$ of non-linear type $A$. It results in an address $l$ paired with an assertion $(l \mapsto_A m)$.
      \item $\textbf{free} : (A : U) \rightarrow F l:\mathbb{N}.F m:A.(l \mapsto_A m) \rightarrow \unit$ \\
      $\free$ is a function that de-allocates the memory cell at location $l$ which is currently in use. This can be seen in its input $F l:\mathbb{N}.F m:A.(l \mapsto_A m)$, where the assertion component $(l \mapsto_A m)$ is a proof that location $l$ is currently allocated.
      \item $\textbf{get} : (A : U) \rightarrow (l : \mathbb{N}) \rightarrow (m : A) \rightarrow (l \mapsto_A m) \rightarrow F \_: (\Sigma x. x =_A m). (l \mapsto_A m)$ \\
      $\get$ is a function that retrieves the term stored at address $l$ when given a proof $(l \mapsto_A m)$ that there is indeed a term $m$ currently stored there. The type of $\get$'s output is the rather complicated $F \_: (\Sigma x. x =_A m). (l \mapsto_A m)$. This not only returns the term stored at $l$, but also a proof that the returned term is exactly the same as what $(l \mapsto_A m)$ claimed to have stored. The proof $(l \mapsto_A m)$ is returned unchanged as well.
      \item $\textbf{set} : (A\ B : U) \rightarrow (l : \mathbb{N}) \rightarrow (m : A) \rightarrow (l \mapsto_A m) \multimap (n : B) \multimap (l \mapsto_B n)$ \\
      $\set$ has the most complicated type of all the axioms presented here. Intuitively, if address $l$ is allocated and storing some term $m$ of type $A$, $\set$ may put term $n$ of type $B$ into the the store at address $l$, overwriting the original term. This is the so called "strong update" operation. The consumption of $(l \mapsto_A m)$ and return of $(l \mapsto_B n)$ accurately characterize this update process.
    \end{itemize}
  \end{definition}
  Due to the fact that $(l \mapsto_A m)$ is a linear type, proofs of this type can not be duplicated and may only be consumed by other state axioms. This is the same non-sharing principle that empowers Separation Logic. Unlike Separation Logic, $(l \mapsto_A m)$ itself is a first-class object in CLC, meaning that it can be the result of computation (large eliminations), input to computations, stored in data structures, etc.

@@

  \subsection{Session Reasoning}
  The inclusion of dependent types and linear types in CLC makes it extremely expressive when encoding communication protocols. By indexing each channel with a protocol, CLC can enforce communication on a channel to strictly adhere to its specified protocol.

  In Figure \ref{session}, we declare an inductive \texttt{session} type for specifying protocols. Intuitively, the \texttt{session}  forms a stack of possible operations executed in a protocol. The $\SEND$ constructor takes in a non-linear type $A$ and a session $ss$ as its argument, indicating that after sending a message of type $A$, the protocol progresses to stage $ss$. Likewise, the $\RECV$ constructor takes in a non-linear type $A$ and a session $ss$ as its argument, indicating that after receiving a message of type $A$, the protocol progresses to stage $ss$. Finally, the $\END$ constructor indicates that the protocol has finished and communication is over.

  \begin{figure}[h]
  \caption{Inductive Session Type}
  \begin{lstlisting}
  Inductive session : U :=
  | SEND : U -> session -> session
  | RECV : U -> session -> session
  | END : session.
  \end{lstlisting}
  \label{session}
  \Description{}
  \end{figure}

  In Definition \ref{comm}, we postulate five axioms for communication using the $\session$ type. These axioms could be viewed as an interface to trusted communication libraries.

  \begin{definition}
    Communication axioms.
    \begin{itemize}
      \item $\textbf{channel} : \session \rightarrow L$ \\
      The $\channel$ axiom is a type constructor. It is used to form the types of communication channels that obey the protocol specified by its session.
      \item $\textbf{open} : (ss : \session) \rightarrow \channel\ ss$ \\
      When given a protocol, $\open$ creates a channel indexed by this protocol.
      \item $\textbf{close} : \channel\ \END \rightarrow \unit$ \\
      Once all communications specified on a given channel has finished, $\close$ will close and reclaim the channel.
      \item $\textbf{send} : (A : U) \rightarrow A \rightarrow (ss : \session) \rightarrow$ \\
      \phantom{send : (A :)}
      $\channel\ (\SEND\ A\ ss) \rightarrow \channel\ ss$ \\
      After a channel has reached the point in its protocol where a message of type $A$ should be sent, $\send$ may send a term of type $A$ on this channel, causing the channel to progress to the next stage of its protocol.
      \item $\textbf{recv} : (A : U) \rightarrow (ss : \session) \rightarrow $ \\
      \phantom{recv : (A :)}
      $\channel\ (\RECV\ A\ ss) \rightarrow F \_:A.\channel\ ss$
      After a channel has reached the point in its protocol where a message of type $A$ should be received, $\recv$ may receive such a term of type $A$ from the channel, causing the channel to progress to the next stage of its protocol.
    \end{itemize}
    \label{comm}
  \end{definition}

  The important detail to notice here is that each channel created by $\open$ is of linear type $(\channel\ ss)$ where $ss$ is some arbitrary protocol, hence must be used exactly once. However, the $\send$ and $\recv$ operations return back channels after consuming them, progressing their protocols one stage forward. Coupled with the fact that the $\close$ operation is only able to close channels whose protocols have ended, this forces channels to communicate strictly by their specified protocols.

  Dependent types also allows for more precise specifications of protocols themselves. A simple example would be the specification of message sizes. Figure \ref{communication} shows such an example. The inductive type \texttt{ilist} is a length indexed list. Using \texttt{ilist}, we construct a protocol \texttt{ss} that first sends two natural numbers, then receives eight natural numbers and finally sends back a boolean acknowledgement.

  \begin{figure}[h]
  \caption{Sized Communication}
  \begin{lstlisting}
  Inductive ilist (A : U) : Nat -> U :=
  | nil  : ilist A 0
  | cons : A -> (n : Nat) -> 
              ilist A n -> ilist A (S n).

  let ss := 
    SEND (ilist Nat 2) 
      (RECV (ilist Nat 8) (SEND Bool END)) 
  in
  let send_msg : ilist Nat 2 := 
    (cons 1 _ (cons 2 _ nil)) 
  in
  let ch := open ss in
  let ch := send _ send_msg _ ch in
  let [ recv_msg, ch ] := recv _ _ ch in
  let ch := send _ true _ ch in
  close ch            
  \end{lstlisting}
  \label{communication}
  \Description{}
  \end{figure}
